\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-21 jasa/paper-jasa.tex
Continued

% Thus far we have shown that the target likelihood ratio
% $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be
% reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the
% classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of
% $p(x|\theta_0)/p(x|\theta_1)$. We now generalize from the ratio of two simple
% hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are
% continuous model parameters. We postpone the practicalities of training the
% classifier and estimating the density to Section~\ref{S:classifier} and continue
% in the likelihood-free setting with idealized classifiers and their densities.
%
% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$
% for the observed data and never refer back to the original features $x_e$. In
% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
% \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
%
% The critical observation is that  if we postpone the evaluation of the
% classifier to the stage of evaluating the enveloping likelihood ratio, then we
% can identify the value of the parameters  that are being compared in the
% likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
%
% \begin{equation}\label{eq:embedding}
% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
% \end{equation}
%
% This is equivalent to approximating the likelihood function for $\theta_0$  when
% $\theta_1$ is held fixed.

@@

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_0^C$, the generalized likelihood ratio\footnote{Also known as
the profile likelihood ratio.} test is commonly used
\begin{equation}
\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
\end{equation}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

Denote the maximum likelihood estimator
\begin{equation}\label{eq:mle}
\hat{\theta} = \argmax_\theta  p(D | \theta)
\end{equation}
and the conditional maximum likelihood estimator
\begin{equation}\label{eq:cmle}
\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p(D | \theta) \; .
\end{equation}

It is not obvious that if we are working with the distributions $p(s|\theta)$
(for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the
same estimators. Fortunately, there is a construction based on $p(s|\theta)$
that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$
for some fixed value of $\theta_1$. This allows us to use Theorem~1 toreformulate the maximum likelihood estimate

\begin{equation}\label{eq:mle_withs}
\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} =
\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $p(s(x_e; \theta, \theta_1)
|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

\subsection{Training a parametrized classifier}

In order to provide parameter inference in the likelihood-free setting, we must
train a family of  classifiers parametrized by $\theta_0$ and $\theta_1$, the
parameters associated to the null and alternate hypotheses, respectively. While
this could be done independently for all $\theta_0$ and $\theta_1$, it is
desirable and convenient to have a smooth evolution of the classification score
as a function of the parameters. Thus, we anticipate a single learning stage
based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$.
Somewhat unusually, the unknown values of the parameters are taken as input to
the classifier; their values will be specified via the enveloping (generalized)
likelihood ratio of Eq.~\ref{eq:enveloping}. We denote the learned family of
classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based
roughly on the following algorithmic flow.
% \begin{algorithm}[ht]
% \caption{Training of the parametrized classifier.}\label{alg:training}
% \begin{algorithmic}
% \STATE initialize trainingData to \{\}
% \FOR{ $\theta_0$ in $\Theta$ }
% 	\FOR{ $\theta_1$ in $\Theta$ }
% 		\STATE generate $x_i^0 \sim p(x|\theta_0)$
% 		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
% 		\STATE generate $x_i^1 \sim p(x|\theta_1)$
% 		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
% 	\ENDFOR
% \ENDFOR
% \STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
% \end{algorithmic}
% \end{algorithm}%\vspace{-1.5em}

While the function $p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
the expected squared loss based on training data produced according to
Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0
\ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world
classifier with finite capacity. This is left as an area for future work.

\subsection{Practicalities of embedding the classifier into the likelihood}

@@

There is a mild technical challenge in embedding the classifier into the
likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to
pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features
$x_e$. In the parametrized setting it is not possible to pre-compute $\hat
s(x_e; \theta_0, \theta_1)$ since one does not know the true values of
$\theta_0$ and $\theta_1$. Thus one must implement the embedding of the
classifier as in Eq.~\ref{eq:embedding}.  A concrete realization of this has
been performed for probability models implemented with the \texttt{RooFit}
probabilistic programing language and  classifiers implemented with
\texttt{scikit-learn} and
\texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

@@

In HEP we are often searching for some class of events, generically referred to
as \textit{signal}, in the presence of a separate class of \textit{background}
events.  For each event we measure some quantities $x$ that have corresponding
distributions $p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where
$\nu$ are nuisance parameters describing uncertainties in the underlying physics
prediction or response of the measurement device. The total model is a mixture
of the signal and background, and $\mu$ is the mixture coefficient associate to
the signal component.
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where
the null corresponds to $\mu=0$, and the generalized likelihood ratio (profiling
over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test
statistic~\citep{Cowan:2010js}. This approach was used for the discovery of the
Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.

\subsection{Typical usage pattern}

@@

\begin{figure}[htbp]
\begin{center}
 \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
 \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT). Right: the corresponding ROC curve (right) for this and other classifiers. (Figures taken from TMVA manual.)}
\label{fig:tmva}
\end{center}
\end{figure}

@@

\subsection{Comments on typical usage of machine learning in HEP}

@@

\subsection{A more powerful  approach}

@@

\subsection{Decomposing tests between mixture models into their components}

@@

\subsection{Measuring particle properties}

@@

\section{Related works}\label{S:Related}

@@

\section{Conclusions}

>>>>

\glnote{Needs some work.}

% Thus far we have shown that the target likelihood ratio
% $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be
% reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the
% classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of
% $p(x|\theta_0)/p(x|\theta_1)$. We now generalize from the ratio of two simple
% hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are
% continuous model parameters. We postpone the practicalities of training the
% classifier and estimating the density to Section~\ref{S:classifier} and continue
% in the likelihood-free setting with idealized classifiers and their densities.

Thus far we have shown that the target likelihood ratio
$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
$\mathbf{x}$ can be reproduced via the univariate densities
$p(s(\mathbf{x};\theta_0,\theta_1)|\theta_0)/p(s(\mathbf{x};\theta_0,\theta_1)|\theta_1)$
if the reduction $s(\mathbf{x}|\theta_0, \theta_1)$ is monotonic with
$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$. We now generalize from the
ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the
case of composite hypothesis testing where $\theta$ are continuous model
parameters.
%
% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$
% for the observed data and never refer back to the original features $x_e$. In
% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
% \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
%
% The critical observation is that  if we postpone the evaluation of the
% classifier to the stage of evaluating the enveloping likelihood ratio, then we
% can identify the value of the parameters  that are being compared in the
% likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
%
% \begin{equation}\label{eq:embedding}
% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
% \end{equation}
%
% This is equivalent to approximating the likelihood function for $\theta_0$  when
% $\theta_1$ is held fixed.

@@

In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
$\theta \in \Theta_0^C$, the generalized likelihood ratio test, also known as
the profile likelihood ratio test, is commonly used
\begin{equation}\label{eqn:generalized-lr}
\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
\end{equation}
\glnote{For the denominator, shouldnt we have $\theta \in \Theta_0^C$ instead?}
This generalized likelihood ratio can be used both for hypothesis tests in the
presence of nuisance parameters or to create confidence intervals with or
without nuisance parameters.  Often, the parameter vector is broken into two
components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
parameters of interest while the $\nu$ components are considered nuisance
parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
$\mu$ fixed.

Let denote the maximum likelihood estimator
\begin{equation}\label{eq:mle}
\hat{\theta} = \argmax_\theta  p({\cal D} | \theta) \; .
\end{equation}
% and the conditional maximum likelihood estimator
% \begin{equation}\label{eq:cmle}
% \hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p({\cal D} | \theta) \; .
% \end{equation}
It is not obvious that if we are working with the distributions $p(s|\theta)$
(for some particular $s(\mathbf{x}; \theta_0, \theta_1)$ comparison) that we can find the
same estimators. Fortunately, there is a construction based on $p(s|\theta)$
that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
the value that maximizes the likelihood ratio with respect to $p({\cal
D}|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use
Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood estimate:
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \; .
\end{align}
Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

\subsection{Parametrized classifier}

\glnote{Needs to be clarified.}
\glnote{Do we put back the algorithm snippet? I think this could be directly
explained in words.}

In order to provide parameter inference in the likelihood-free setting, we must
train a family of classifiers parametrized by $\theta_0$ and $\theta_1$, the
parameters associated to the null and alternate hypotheses, respectively. While
this could be done independently for all $\theta_0$ and $\theta_1$, it is
desirable and convenient to have a smooth evolution of the classification score
as a function of the parameters. Thus, we anticipate a single learning stage
based on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$.
Somewhat unusually, the unknown values of the parameters are taken as input to
the classifier; their values will be specified via the enveloping (generalized)
likelihood ratio of Eqn.~\ref{eqn:generalized-lr}. We denote the learned family of
classifiers $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$, and anticipate the training based
roughly on the following algorithmic flow.

% \begin{algorithm}[ht]
% \caption{Training of the parametrized classifier.}\label{alg:training}
% \begin{algorithmic}
% \STATE initialize trainingData to \{\}
% \FOR{ $\theta_0$ in $\Theta$ }
% 	\FOR{ $\theta_1$ in $\Theta$ }
% 		\STATE generate $x_i^0 \sim p(x|\theta_0)$
% 		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
% 		\STATE generate $x_i^1 \sim p(x|\theta_1)$
% 		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
% 	\ENDFOR
% \ENDFOR
% \STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
% \end{algorithmic}
% \end{algorithm}%\vspace{-1.5em}

While the function $p(\mathbf{x}|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
the expected squared loss based on training data produced according to
Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0
\ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world
classifier with finite capacity. This is left as an area for future work.

\subsection{Parametrized calibration}

\glnote{Needs to be clarified as well.}

@@

% There is a mild technical challenge in embedding the classifier into the
% likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to
% pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features
% $x_e$. In the parametrized setting it is not possible to pre-compute $\hat
% s(x_e; \theta_0, \theta_1)$ since one does not know the true values of
% $\theta_0$ and $\theta_1$. Thus one must implement the embedding of the
% classifier as in Eq.~\ref{eq:embedding}.
%  A concrete realization of this has
% been performed for probability models implemented with the \texttt{RooFit}
% probabilistic programing language and  classifiers implemented with
% \texttt{scikit-learn} and
% \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

@@

\subsection{High energy physics}

\glnote{Add a concrete HEP example where we illustrate and compare
the method for several classifiers and calibration algorithms.}
\glnote{Notations need to be adapted.}
\glnote{Rethink the outline? Start with comparison of mixtures (5.1.4) and then
specialize to background vs background+signal, and thereby discuss that we can
focus the capacity of the classifier.}

In high energy physics, we are often searching for some class of events,
generically referred to as \textit{signal}, in the presence of a separate class
of \textit{background} events.  For each event we measure some quantities $x$
that have corresponding distributions $p_b(x|\nu)$ for background and
$p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
uncertainties in the underlying physics prediction or response of the
measurement device. The total model is a mixture of the signal and background,
and $\mu$ is the mixture coefficient associate to the signal component.
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where
the null corresponds to $\mu=0$, and the generalized likelihood ratio (profiling
over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test
statistic~\citep{Cowan:2010js}. This approach was used for the discovery of the
Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.

\subsubsection{Typical usage pattern}

@@

\begin{figure}[t]
\begin{center}
 \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
 \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT). Right: the corresponding ROC curve (right) for this and other classifiers. (Figures taken from TMVA manual.)}
\label{fig:tmva}
\end{center}
\end{figure}

@@

\subsubsection{Comments on typical usage of machine learning in HEP}

@@

\subsubsection{A more powerful  approach}

@@

\subsubsection{Decomposing tests between mixture models into their components}

@@

\subsection{Likelihood-free inference}

\glnote{Rework this section, add a concrete example and compare with other
likelihood-free inference algorithms.}

@@

\section{Related works}
\label{sec:related}

\glnote{Add related works on density ratio estimation.}
\glnote{Remove others that are less directly related.}

@@

\section{Conclusions}
\label{sec:conclusions}
